Division and Slope Factorization of p-Adic Polynomials

### 2016-04-29 polynomials/slopefact.tex
Revision of the slope factorization paper

\permission{%
}

@@

\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulation -- \emph{Algebraic Algorithms}}
\terms{Algorithms, Theory}
\keywords{Algorithms, $p$-adic precision, Newton polygon, factorization}

@@

\noindent
{\bf Main Results.}
%
Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
\begin{align*}
A_{i+1} &= A_i + (V_i P \mod A_i) \\
B_{i+1} &= P \div A_{i+1} \\
V_{i+1} &= (2V_i - V_i^2B_{i+1}) \mod A_{i+1}.
\end{align*}
Our main result is Theorem \ref{theo:slope-factor},
which states that the sequence $(A_i)$ converges quadratically
to a divisor of $P$. This provides a quasi-optimal simple-to-implement 
algorithm for computing slope factorizations.
We moreover carry out a careful study of the precision and, applying
a strategy coming from~\cite{caruso-roe-vaccon:14a}, we end up with
an algorithm that outputs optimal results regarding to accuracy.

@@

Finally, in Section \ref{sec:slope_fac} we describe
our slope factorization algorithm, which is based on
a Newton iteration.  Unlike other algorithms for
slope factorization, ours does not require working
with fractional exponents. % \todo{ citation?}. 
In Theorem \ref{theo:slope-factor} we define a sequence
of polynomials that will converge to the factors
determined by an extremal point in the Newton
polygon.  We then discuss the precision behavior
of the algorithm.

@@

In what follows, the notation $K[X]$ refers to the ring of univariate 
polynomials with coefficients in $K$. The subspace of polynomials of 
degree at most $n$ (resp. exactly $n$) is denoted by $K_{\leq n}[X]$ 
(resp. $K_{=n}[X]$).

@@

\noindent
{\bf Flat precision.}
The simplest method for tracking the precision of a polynomial is
to record each coefficient modulo a fixed power of $p$.  While
easy to analyze and implement, this method suffers when
applied to polynomials whose Newton polygons are far from flat.

@@

\begin{deftn}
A \emph{Newton function of degree $n$} is a convex function 
$\varphi : [0,n] \to \R \cup \{+\infty\}$ which is piecewise affine, 
which takes a finite value at $n$ and whose epigraph $\Epi(\varphi)$ 
have extremal points with integral abscissa.
\end{deftn}

@@

\begin{deftn}
\label{def:nondeg}
Let $P = P_\app + O(\varphi_P)$. We say that the Newton precision 
$O(\varphi_P)$ on $P$ is \emph{nondegenerate} if $\varphi_P \geq 
\NF(P_\app)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
$\NP(P_\app)$.
\end{deftn}

@@

\noindent
{\bf Newton precision.}
We first analyze the precision behavior of Euclidean division
in the Newton model.  Concretely, we pick $A, B \in K[X]$ two polynomials which 
are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
$$A = A_\app + O(\varphi_A)
\quad \text{and} \quad
B = B_\app + O(\varphi_B)$$
Here $A_\app$ and $B_\app$ are some approximations of $A$ and $B$ respectively 
and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
$\deg A$ and $\deg B$ respectively.
We are interested in determining the precision on $A \mod B$ and $A
\div B$. The following proposition gives a theoretical answer under
mild assumptions.

@@

\noindent
{\bf Lattice precision.}
We now move to lattice precision. We pick $A$ and $B$ two polynomials 
of respective degree $n$ and $d$ and assume that they are known at 
precision $O(H_A)$ and $O(H_B)$ respectively:
$$A = A_\app + O(H_A)
\quad \text{and} \quad
B = B_\app + O(H_B).$$
where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
According to the results of \cite{caruso-roe-vaccon:14a}, in order to determine the 
precision on $A \div B$ and $A \mod B$, we need to compute the 
differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
\mapsto X \mod Y$ at the point $(A_\app, B_\app)$. Writing $Q_\app = A_\app \div
B_\app$ and $R_\app = A_\app \mod B_\app$, this can be done by
expanding the relation:
$$A_\app + dA = (B_\app + dB) (Q_\app + dQ) + (R_\app + dR)$$
and neglecting the terms of order $\geq 2$. We get this way 
$dA = B_\app dQ + Q_\app dB + dR$
meaning that $dQ$ and $dR$ appears respectively as the quotient and
the remainder of the Euclidean division of $dX = dA - Q_\app dB$ by $B_\app$.

@@

\noindent $\bullet$
in the case of jagged and Newton precision, the precision on the output 
may be written into the form $\sum_{i=0}^4 O(2^{N_i}) \: X^i$ and 
$G = \sum_{i=0}^4 (N_i - N)$;

\noindent $\bullet$
in the case of lattice precision, the precision on the output is a 
lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
(The term $G_d$ corresponds to diffused digits according to 
\cite[Definition~2.3]{caruso-roe-vaccon:15a}.)

@@

@@

\begin{comment}
\subsubsection{Tristan's proof}

\begin{deftn}
Notation:
\[P=-d \lambda_0+d \slp(\lambda_0) + l_1 \slp (\lambda_1)+\RS \]
($\RS$ for remaining slopes)

\begin{tikzpicture}
\draw[->] (0,-1) -- (0,2);

\draw[->] (-0.05,0) -- (4,0);

\node[left] at (0,0) {$0$};

\draw (1.5,-.1) -- (1.5,.1);
\node[above] at (1.5,0) {$d$};

\draw (3,-.1) -- (3,.1);
\node[above] at (2.8,0) {$d+l_1$};

\node[left] at (0,1.5) {$-d \lambda_0$};

\draw (0,1.5)-- (1.5,0) -- (3,-.375)-- (3.5,.625);
\node[below] at (.75,0.5) {$\lambda_0$};
\node[below] at (2.25,-0.25) {$\lambda_1$};
\node[right] at (3.5,0.25) {$\RS$};
\end{tikzpicture}
\end{deftn}

{\bf Initial Newton polygons}

\begin{lem}
$\NP(R_0)=\NP(S_0)$:
\[ \NP (R_0)=\NP(S_0)=(\lambda_1-\lambda_0)- d \lambda_0+(d-1)\slp( \lambda_0). \]

\begin{tikzpicture}
\draw[->] (0,-.5) -- (0,2.5);

\draw[->] (-0.05,0) -- (2,0);

\draw[->] (-0.05,1.75) -- (.05,1.75);
\node[left] at (0,1.75) {$(\lambda_1-\lambda_0)$};
\node[left] at (-.4,1.4) {$-d \lambda_0$};

\node[left] at (0,0) {$0$};

\draw (1,-.1) -- (1,.1);
\node[below, right] at (1,-.2) {$d-1$};

\draw (0,1.75) -- (1,.75) -- (1,2.5);
\node[below] at (.5,1.3) {$\lambda_0$};
\end{tikzpicture}

$\NP (1-V_0B_0 \% A_0)$:
\begin{align*}
\NP ((1-V_0B_0)\% A_0) &\geq (\lambda_1-\lambda_0)+1\slp( \lambda_0)+(d-1) \slp(\lambda_1), \\
&\geq (\lambda_1-\lambda_0)+d\slp( \lambda_0). \\
\end{align*} 
\begin{tikzpicture}
\draw[->] (0,-1) -- (0,1);

\draw[->] (-0.05,0) -- (2.5,0);

\node[left] at (0,0) {$0$};

\node[above] at (1.5,0) {$d-1$};

\draw (.5,-.1) -- (.5,.1);
\node[above] at (.5,0) {$1$};

\draw (-.1,.25) -- (.1,.25);
\node[left] at (0,.25) {$(\lambda_1-\lambda_0)$};

\draw (0,.25) -- (.5,-.25) -- (1,-.375)-- (1,1.5);
\draw (0,.25) -- (.5,-.25) -- (1,-.75)-- (1,1.5);

\node[below] at (.25,-0.1) {$\lambda_0$};
\node[below] at (.75,0.1) {$\lambda_1$};
\end{tikzpicture}

\label{lem:initial_NP}
\end{lem}
\begin{proof}
Clear by Theorem \ref{theo:EDivisionNP}, and the fact that $coeff(P,d)=1.$
\end{proof}

{\bf Unchanged Newton polygons}

\begin{lem} \label{lem_unchanged_NP}
$\NP (V_i)$ for $i>0$ ($V_0=1$):
\[\NP(V_i)=0+(d-1)\slp( \lambda_1). \]

\begin{tikzpicture}
\draw[->] (0,-1) -- (0,1);

\draw[->] (-0.05,0) -- (2,0);

\node[left] at (0,0) {$0$};

\node[below, right] at (1,-.2) {$d-1$};

\draw (0,0) -- (1,-.25) -- (1,1);
\node[below] at (.5,-0.25) {$\lambda_1$};
\end{tikzpicture}

$\NP (B_i)$:
\[\NP(B_i)=0+(l_1)\slp( \lambda_1)+\RS. \]

\begin{tikzpicture}
\draw[->] (0,-1) -- (0,1);

\draw[->] (-0.05,0) -- (2.5,0);

\node[left] at (0,0) {$0$};

\draw (1.5,-.1) -- (1.5,.1);
\node[above] at (1.5,0) {$l_1$};

\draw (0,0) -- (1.5,-.375) -- (2,.625);
\node[below] at (.5,-0.25) {$\lambda_1$};
\node[below] at (2,-0.25) {$\lambda_2$};
\end{tikzpicture}

$\NP (A_i)$:
\[\NP(A_i)=-d \lambda_0+d \slp( \lambda_0). \]

\begin{tikzpicture}
\draw[->] (0,-1) -- (0,2);

\draw[->] (-0.05,0) -- (2,0);

\node[left] at (0,0) {$0$};

\draw (1.5,-.1) -- (1.5,.1);
\node[above] at (1.5,0) {$d$};

\node[left] at (0,1.5) {$-d \lambda_0$};

\draw (0,1.5)-- (1.5,0) ;
\node[below] at (.75,0.5) {$\lambda_0$};
\end{tikzpicture}
\end{lem}

{\bf Newton polygons inside the loop}

\begin{lem} \label{lem:iteration_inside_the_loop}
\begin{align*}
\NP (R_i), \NP(S_i) & \geq -\lambda_0+2^i(\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0), \\
\NP (1-V_i B_i) & \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0),\\
\end{align*}
\end{lem}

We prove the last two lemmas by induction.

\begin{proof}
The case $i=0$ is clear thanks to Theorem \ref{theo:EDivisionNP} and Lemma \ref{lem:initial_NP}.

Now, we assume that for some $i \geq 0$ the formulae defined in the Lemmae \ref{lem_unchanged_NP} and \ref{lem:iteration_inside_the_loop} are satisfied. 
We prove them for $i+1$.

Firstly, thanks to the minoration on $\NP (R_i)$ and the fact that $R_i=A_{i+1}-A_i$, we have $\NP (A_{i+1}= \NP (A_i).$
Since $B_{i+1} = P \div A_{i+1},$ we have $\NP (B_{i+1}) = \NP (B_i).$
We apply \eqref{eqdef:Biminus} to obtain 
\[\NP (B_i-B_{i+1})  \geq 2^i (\lambda_1-\lambda_0)+(\lambda_1-\lambda_0)+(l_1-1) \slp (\lambda_1)+\RS. \]
Then, we can apply \eqref{eqdef:Siminus} to obtain 
\[\NP (S_i-S_{i+1})  \geq -\lambda_0 d+2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]

As a consequence, we get that $\NP (S_{i+1}) \geq \NP (S_i)$ (pointwise).

Now, we can apply \eqref{eqdef:Viminus} to obtain that 
\[\NP (V_i-V_{i+1}) \geq 2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]
As a consequence, we get that $\NP(V_i)=\NP(V_{i+1}).$
Similarly, we use \eqref{eqdef:Qiminus} to obtain that
\[\NP (1-Q_i)  \geq  2^i (\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0).\]

This is enough to conclude on the Newton polygon of $R_{i+1}.$
Indeed, thanks to \eqref{eq:Riplus}, we obtain that 
\[\NP (R_i) \geq -\lambda_0+2^i(\lambda_1-\lambda_0)+(d-1) \slp (\lambda_0). \]
We can proceed similarly for $1-V_{i+1} B_{i+1}$ using \eqref{eq:ViBiplus}.

Finally, we can apply \eqref{eqdef:Si2} for $i+1,$
\begin{align*}
S_{i+1} & = (B_{i+1} R_{i+1} + (1-V_{i+1} B_{i+1}) S_{i} \\
& + (1-V_{i+1} B_{i+1})(S_{i+1}-S_{i})) \% A_{i+1},
\end{align*}
to obtain the desired minoration on $\NP (S_{i+1}).$ This concludes the proof.
\end{proof}

\subsubsection{Xavier's proof}
\end{comment}

@@

Let $P \in K_n[X]$ and $d$ be the abscissa of an extremal point of 
$\NP(P)$. Previously (\emph{cf} Theorem~\ref{theo:slope-factor}), we 
have defined a sequence $(A_i, V_i)$ converging to $(A,V)$ where $A$ is 
a factor of $P$ whose Newton function is $\NF(P)_{|[0,d]}$ and $V$ is the inverse of $B = P/A$ modulo $A$.
We now assume that $P$ is known up to some finite precision: $P = P_\app + 
O(\hspace{0.5mm}\cdots)$ where the object inside the $O$ depends on the 
chosen precision model. We address the two following questions:
(1)~what is the precision on the factor $A$, and
(2)~how can one compute in practice $A$ at this precision?

In the sequel, it will be convenient to use a different normalization on 
$A$ and $B$: if $a_d$ is the coefficient of $P$ of degree $d$, we set
$A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We shall also 
always assume that $P$ is monic in the sense that its leading
coefficient is \emph{exactly}~$1$; the precision datum on $P$ then 
only concerns the coefficients up to degree $n{-}1$. Similarly, noting
that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
precision datum up to degree $d{-}1$ and $n{-}d{-}1$ respectively.

@@

\noindent
{\bf Newton precision.}
We assume that the precision on the input $P$ has the shape 
$O(\varphi_P)$ where $\varphi_P$ is a Newton function of degree 
$n{-}1$. From now on, we assume that the precision $O(\varphi_P)$ 
is nondegenerate in the sense of Definition~\ref{def:nondeg}. This
ensures in particular that the Newton polygon of $P$ is well defined. 
We import the notations $\varphi$, $b_\varphi$ and $b_0$ from \S 
\ref{ssec:Newtoniter} and refer to 
Eqs.~\eqref{eqdef:functionphi}--\eqref{eqdef:b0} for the definitions.

@@

\noindent
Obviously \textbf{(H)} gives an answer to the first question we have 
raised above: the precision on the couple $A^{(1)}$ is the lattice 
$H_{A^{(1)}}$ defined as projection on the first component of 
$F'(P_\app)(H_P)$. It turns out that it can be computed explicitely as 
shown by the next proposition.

@@

\begin{comment}
\noindent
{\bf Numerical experimentations.} 
%
We have implemented the following scheme to illustrate the behaviour of 
precision for slope factorization over $\mathbb{Q}_2 [X].$
We start with random polynomials $A$ and $B$ with given minoration on 
their Newton polygons, assuring they have no common slope.
We compute $P=AB$ and then $A^{(1)}$, and compare the obtained 
precision on $A^{(1)}$ (jagged) with the estimation given using 
$\psi_{A^{(1)}}$ and Proposition \ref{prop:precA1} 
assuming~\textbf{(H)}. The mean amount of lost digits for $1000$ 
experiments is given in the following array. Two settings are possible 
for the initial $A$ and $B$ : relative or absolute precision equal to 
$100$ for their coefficients.

{\small%
\noindent \hfill%
\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|c|c|c|c|c|}%
\hline
\multirow{2}{*}{Polynomials} & 
\multirow{2}{*}{\hspace{1mm}Precision\hspace{1mm}} & 
\multicolumn{3}{c|}{Gain of precision} \\
\cline{3-5}
& &Jagged & Newton & Lattice \\
\hline
\multirow{2}{*}{%
\begin{tikzpicture}[xscale=0.11, yscale=0.05]
\draw[transparent] (-2,11) rectangle (15,11);
\draw[->] (0,-.5) -- (0,10.5);
\draw[->] (-0.05,0) -- (13.5,0);
\draw (6,-.75) -- (6,.75);
\node[scale=0.7,below] at (6,0) {$d$};
\draw (0,5) -- (2,-2) -- (6,0) -- (10,3) -- (13,10);
\draw[dotted] (6,0) -- (6,10);
\node at (3.5,5) {$A$};
\node at (8.5,5) {$B$};
\end{tikzpicture}}
& absolute & $-14.5$ & $-14.7$ & $0.0$ \\
& relative & $\phantom{0}{-}1.5$ & $\phantom{0}{-}3.6$ & $0.0$ \\
\hline
\multirow{2}{*}{\begin{tikzpicture}[xscale=0.11, yscale=0.05]
\draw[transparent] (-2,11) rectangle (15,11);
\draw[->] (0,-.5) -- (0,10.5);
\draw[->] (-0.05,0) -- (13.5,0);
\draw (6,-.75) -- (6,.75);
\node[scale=0.7,below] at (6,0) {$d$};
\draw (0,5) -- (6,0) -- (11,-2) -- (13,10);
\draw[dotted] (6,0) -- (6,10);
\node at (3.5,5) {$A$};
\node at (8.5,5) {$B$};
\end{tikzpicture}}
& absolute & $\phantom{-0}0.0$ & $\phantom{-0}0.0$ & $0.0$ \\
& relative & $\phantom{0}{-}0.3$ & $\phantom{0}{-}1.2$ & $0.0$ \\
\hline
\end{tabular}
\hfill \null}
\end{comment}

\bibliographystyle{plain}
\bibliography{muldivfact}

>>>>

\CopyrightYear{2016}
\setcopyright{licensedothergov}
\conferenceinfo{ISSAC'16,}{July 19--22, 2016, Waterloo, ON, Canada} \isbn{978-1-4503-4380-0/16/07}
\acmPrice{\$15.00}
\doi{http://dx.doi.org/10.1145/2930889.2930897}

\permission{Publication rights licensed to ACM. ACM acknowledges that this contribution was authored or co-authored by an employee, contractor or affiliate of a national government. As such, the Government retains a nonexclusive, royalty-free right to publish or reproduce this article, or to allow others to do so, for Government purposes only.}

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010147.10010148.10010149.10010150</concept_id>
<concept_desc>Computing methodologies~Algebraic algorithms</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\vspace{-1mm}
\ccsdesc[500]{Computing methodologies~Algebraic algorithms}
\printccsdesc

\vspace{-1.5mm}
\keywords{Algorithms, $p$-adic precision, Newton polygon, factorization}

@@

\noindent
{\bf Main Results.}
%
Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
has an extremal point at abscissa $d$.
Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
\begin{align*}
A_{i+1} &= A_i + (V_i P \mod A_i) \\
B_{i+1} &= P \div A_{i+1} \\
V_{i+1} &= (2V_i - V_i^2B_{i+1}) \mod A_{i+1}.
\end{align*}
where $S \div T$ and $S \mod T$ denotes the quotient and the
remainder in the Euclidean division of $S$ by $T$ respectively.
Our main result is Theorem \ref{theo:slope-factor},
which states that the sequence $(A_i)$ converges quadratically
to a divisor of $P$. This provides a quasi-optimal simple-to-implement 
algorithm for computing slope factorizations.
We moreover carry out a careful study of the precision and, applying
a strategy coming from~\cite{caruso-roe-vaccon:14a}, we end up with
an algorithm that outputs optimal results regarding to accuracy.

@@

Finally, in Section \ref{sec:slope_fac} we describe
our slope factorization algorithm, which is based on
a Newton iteration.  Unlike other algorithms for
slope factorization, ours does not require working
with fractional exponents (compare for instance with
\cite[\S~6]{poteaux-rybowicz}).
In Theorem \ref{theo:slope-factor} we define a sequence
of polynomials that will converge to the factors
determined by an extremal point in the Newton
polygon.  We then discuss the precision behavior
of the algorithm.

@@

In what follows, the notation $K[X]$ refers to the ring of univariate 
polynomials with coefficients in $K$. The subspace of polynomials of 
degree at most $n$ (resp. exactly $n$) is denoted by $K_{\leq n}[X]$ 
(resp. $K_n[X]$).

@@

\noindent
{\bf Flat precision.}
The simplest method for tracking the precision of a polynomial is
to record each coefficient modulo a fixed power of $\pi$.  While
easy to analyze and implement, this method suffers when
applied to polynomials whose Newton polygons are far from flat.

@@

\begin{deftn}
A \emph{Newton function of degree $n$} is a convex function 
$\varphi : [0,n] \to \R \cup \{+\infty\}$ which is piecewise affine, 
which takes a finite value at $n$ and whose epigraph\footnote{Recall
that the epigraph is the region above the graph.} $\Epi(\varphi)$ 
have extremal points with integral abscissa.
\end{deftn}

@@

\begin{deftn}
\label{def:nondeg}
Let $P = P_\app + O(\varphi_P)$. We say that the Newton precision 
$O(\varphi_P)$ of $P$ is \emph{nondegenerate} if $\varphi_P \geq 
\NF(P_\app)$ and $\varphi_P(x) > y$ for all extremal point $(x,y)$ of 
$\NP(P_\app)$.
\end{deftn}

@@

\noindent
{\bf Newton precision.}
We first analyze the precision behavior of Euclidean division
in the Newton model.  Concretely, we pick $A, B \in K[X]$ two polynomials which 
are known at precision $O(\varphi_A)$ and $O(\varphi_B)$ respectively:
$$A = A_\app + O(\varphi_A)
\quad \text{and} \quad
B = B_\app + O(\varphi_B)$$
Here $A_\app$ and $B_\app$ are some approximations of $A$ and $B$ respectively 
and $\varphi_A$ and $\varphi_B$ denotes two Newton functions of degree 
$\deg A$ and $\deg B$ respectively.
We are interested in determining the precision of $A \mod B$ and $A
\div B$. The following proposition gives a theoretical answer under
mild assumptions.

@@

\noindent
{\bf Lattice precision.}
We now move to lattice precision. We pick $A$ and $B$ two polynomials 
of respective degree $n$ and $d$ and assume that they are known at 
precision $O(H_A)$ and $O(H_B)$ respectively:
$$A = A_\app + O(H_A)
\quad \text{and} \quad
B = B_\app + O(H_B).$$
where $H_A \in K_{\leq n}[X]$ and $H_B \in K_{\leq d}[X]$ are lattices. 
According to the results of \cite{caruso-roe-vaccon:14a}, in order to determine the 
precision of $A \div B$ and $A \mod B$, we need to compute the 
differential of the mappings $(X,Y) \mapsto X \div Y$ and $(X,Y)
\mapsto X \mod Y$ at the point $(A_\app, B_\app)$. Writing $Q_\app = A_\app \div
B_\app$ and $R_\app = A_\app \mod B_\app$, this can be done by
expanding the relation:
$$A_\app + dA = (B_\app + dB) (Q_\app + dQ) + (R_\app + dR)$$
and neglecting the terms of order $\geq 2$. We get this way 
$dA = B_\app dQ + Q_\app dB + dR$
meaning that $dQ$ and $dR$ appears respectively as the quotient and
the remainder of the Euclidean division of $dX = dA - Q_\app dB$ by $B_\app$.

@@

\noindent $\bullet$
in the case of jagged and Newton precision, the precision of the output 
may be written into the form $\sum_{i=0}^4 O(2^{N_i}) \: X^i$ and 
$G = \sum_{i=0}^4 (N_i - N)$;

\noindent $\bullet$
in the case of lattice precision, the precision of the output is a 
lattice $H$ and $G$ is the index of $H$ in $2^N \mathcal L$ where 
$\mathcal L = \Z_2[X]/M$ is the standard lattice; in that case, we write 
$G$ as a sum $G_{\text{nd}} + G_{\text{d}}$ where $G_{\text{d}}$ is the 
index of $H$ in the largest lattice $H_0$ contained in $H$ which can be 
generated by elements of the shape $2^{N_i} X^i$ ($0 \leq i \leq 4$). 
(The term $G_d$ corresponds to diffused digits according to 
\cite[Definition~2.3]{caruso-roe-vaccon:15a}.)

@@

\begin{rem}
The iteration formulae of the above theorem already appeared 
in \cite[\S~3.1]{berthomieu:hal-00642075}; nevertheless, as far
as we know, the result of Theorem~\ref{theo:slope-factor} is
new.
\end{rem}

@@

@@

Let $P \in K_n[X]$ and $d$ be the abscissa of an extremal point of 
$\NP(P)$. Previously (\emph{cf} Theorem~\ref{theo:slope-factor}), we 
have defined a sequence $(A_i, V_i)$ converging to $(A,V)$ where $A$ is 
a factor of $P$ whose Newton function is $\NF(P)_{|[0,d]}$ and $V$ is the inverse of $B = P/A$ modulo $A$.
We now assume that $P$ is known up to some finite precision: $P = P_\app + 
O(\hspace{0.5mm}\cdots)$ where the object inside the $O$ depends on the 
chosen precision model. We address the two following questions:
(1)~what is the precision of the factor $A$, and
(2)~how can one compute in practice $A$ at this precision?

In the sequel, it will be convenient to use a different normalization on 
$A$ and $B$: if $a_d$ is the coefficient of $P$ of degree $d$, we set
$A^{(1)} = a_d^{-1} A$ and $B^{(1)} = a_d B$
so that $A^{(1)}$ is monic and $P = A^{(1)} B^{(1)}$. We shall also 
always assume that $P$ is monic in the sense that its leading
coefficient is \emph{exactly}~$1$; the precision datum of $P$ then 
only concerns the coefficients up to degree $n{-}1$. Similarly, noting
that $A^{(1)}$ and $B^{(1)}$ are monic as well, they only carry a
precision datum up to degree $d{-}1$ and $n{-}d{-}1$ respectively.

@@

\noindent
{\bf Newton precision.}
We assume that the precision of the input $P$ has the shape 
$O(\varphi_P)$ where $\varphi_P$ is a Newton function of degree 
$n{-}1$. From now on, we assume that the precision $O(\varphi_P)$ 
is nondegenerate in the sense of Definition~\ref{def:nondeg}. This
ensures in particular that the Newton polygon of $P$ is well defined. 
We import the notations $\varphi$, $b_\varphi$ and $b_0$ from \S 
\ref{ssec:Newtoniter} and refer to 
Eqs.~\eqref{eqdef:functionphi}--\eqref{eqdef:b0} for the definitions.

@@

\noindent
Obviously \textbf{(H)} gives an answer to the first question we have 
raised above: the precision of the couple $A^{(1)}$ is the lattice 
$H_{A^{(1)}}$ defined as projection on the first component of 
$F'(P_\app)(H_P)$. It turns out that it can be computed explicitely as 
shown by the next proposition.

@@

\bibliographystyle{plain}

% \bib, bibdiv, biblist are defined by the amsrefs package.
\begin{bibdiv}
\begin{biblist}
\tiny

\bib{berthomieu-hoeven-lecerf:11a}{article}{
      author={Berthomieu, J{\'e}r{\'e}my},
      author={van~der Hoeven, Joris},
      author={Lecerf, Gr{\'e}goire},
       title={{Relaxed algorithms for $p$-adic numbers}},
        date={2011},
     journal={J. Th{\'e}orie des Nombres des Bordeaux},
      volume={23},
      number={3},
       pages={{541\ndash 577}},
}

\bib{berthomieu:hal-00642075}{article}{
  TITLE = {{Polynomial root finding over local rings and application to error correcting codes}},
  author = {Berthomieu, J{\'e}r{\'e}my},
  author = {Lecerf, Gr{\'e}goire},
  author = {Quintin, Guillaume},
  JOURNAL = {{Applicable Algebra in Engineering, Communication and Computing}},
  PUBLISHER = {{Springer Verlag}},
  VOLUME = {24},
  NUMBER = {6},
  PAGES = {413\ndash 443},
  YEAR = {2013},
  DOI = {10.1007/s00200-013-0200-5},
}

\bib{caruso:16}{unpublished}{
      author={Caruso, Xavier},
       title={Slope factorization of ore polynomials},
        date={2016},
        note={in preparation},
}

\bib{caruso-roe-vaccon:14a}{article}{
      author={Caruso, Xavier},
      author={Vaccon, Tristan},
      author={Roe, David},
       title={Tracking $p$-adic precision},
        date={2014},
     journal={LMS Journal of Computation and Mathematics},
      volume={17 (Special issue A)},
       pages={274\ndash 294},
}

\bib{caruso-roe-vaccon:15a}{inproceedings}{
      author={Caruso, Xavier},
      author={Vaccon, Tristan},
      author={Roe, David},
       title={$p$-adic stability in linear algebra},
        date={2015},
   booktitle={Proceedings of the 2015 acm on international symposium on
  symbolic and algebraic computation},
   publisher={ACM},
     address={New York},
       pages={{101\ndash 108}},
}

\bib{dwork-geratto-sullivan:Gfunctions}{book}{
      author={Dwork, Bernard},
      author={Gerotto, Giovanni},
      author={Sullivan, Francis},
       title={{An introduction to $G$-functions}},
   publisher={Princeton U.P.},
     address={Princeton},
        date={1994},
}

\bib{guardia-montes-nart:08a}{article}{
      author={Gu\`ardia, Jordi},
      author={Montes, Jes\'us},
      author={Nart, Enric},
       title={Newton polygons of higher order in algebraic number theory},
        date={2012},
     journal={Transactions of the AMS},
      volume={364},
      number={1},
       pages={361\ndash 416},
}

\bib{guardia-nart-pauli:12a}{article}{
      author={Guàrdia, Jordi},
      author={Nart, Enric},
      author={Pauli, Sebastian},
       title={Single-factor lifting and factorization of polynomials over local
  fields},
        date={2012},
     journal={Journal of Symbolic Computation},
      volume={47},
      number={11},
       pages={1318 \ndash  1346},
}

\bib{montes:99a}{thesis}{
      author={Montes, Jes\'us},
       title={Pol\'igonos de newton de orden superior y aplicaciones
  aritm\'eticas},
        type={Ph.D. Thesis},
        date={1999},
}

\bib{pauli:10a}{inproceedings}{
      author={Pauli, Sebastian},
       title={{Factoring polynomials over local fields II}},
        date={2010},
   booktitle={Algorithmic number theory, 9th international symposium},
      editor={Hanrot, Guillaume},
      editor={Morain, Fran{\c c}ois},
      editor={Thom\'e, Emmanuel},
     address={Nancy, France},
}

\bib{hoeven:02a}{article}{
      author={van~der Hoeven, Joris},
       title={{Relax, but don't be too lazy}},
        date={2002},
     journal={J. Symbolic Comput.},
      volume={34},
      number={6},
       pages={479\ndash 542},
}

\bib{hoeven:07a}{article}{
      author={van~der Hoeven, Joris},
       title={New algorithms for relaxed multiplication},
        date={2007},
     journal={J. Symbolic Comput.},
      volume={42},
      number={8},
       pages={792\ndash 802},
}

\bib{poteaux-rybowicz}{inproceedings}{
  author    = {Poteaux, Adrien},
  author    = {Rybowicz, Marc},
  title     = {Improving Complexity Bounds for the Computation of Puiseux Series over Finite Fields},
  booktitle = {ISSAC'15},
  pages     = {299--306},
  year      = {2015},
}

\end{biblist}
\end{bibdiv}
